---
title: "Batch processing of HyMi kinematics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Batch processing of HyMi kinematics}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

This vignette demonstrates how to run the batch kinematics pipeline on
a directory of tracking files and associated metadata. The wrapper
function `batch_compute_kinematics()` reads the metadata table,
selects trials marked for inclusion, matches them to `.trk` files, and
runs `compute_kinematics()` for each matched file.

The result is a list of per-trial results that can be combined into a
single data frame for downstream analysis and plotting.

# Data layout

The example assumes the following layout on disk:

- An Excel metadata file with one row per trial
(e.g. `Tabelle_Gesamt_23-4-21.xlsx`).
- A directory containing all `.trk` tracking files for those trials
https://opensourcephysics.github.io/tracker-website/.

Each row in the metadata table contains an individual identifier and
measurement number, which together encode the trial ID used in the
`.trk` filenames.


# Running the batch kinematics wrapper

The main entry point is `batch_compute_kinematics()`. It takes the
directory containing the metadata file (`meta_dir`), the metadata
filename (`meta_file`), the directory with `.trk` files (`trk_dir`),
and a few options controlling the plots produced by
`compute_kinematics()`.

```{r batch-run, eval=FALSE}
library(gothouant)  

root_data_dir <- "~/Dropbox/hymi/v03/data"

res <- batch_compute_kinematics(
  meta_dir    = root_data_dir,
  meta_file   = "Tabelle_Gesamt_23-4-21.xlsx",
  trk_dir     = root_data_dir,
  plotsize    = 1.5,
  scale_plot  = "bodylength",
  out_fig_dir = "~/tmp/gothouant/"
)
```

The function prints the name of each `.trk` file as it is processed. If
`compute_kinematics()` throws an error for a file, the wrapper emits a
warning and returns `NULL` at that position in the result list.

# Building a combined result table

The list `res` contains one element per processed trial. These elements
can be combined row-wise into a single data frame:

```{r bind-res, eval=FALSE}
res_df <- dplyr::bind_rows(res)

dplyr::glimpse(res_df)
```

Each row of `res_df` contains the metadata columns plus list-columns
for derived kinematic summaries, such as:

- `stride_kin`: per-leg and midpoint stride-level measures.
- `leg_times`: per-leg swing, contact, and cycle timings.
- `tarsus_summary`: body-frame tarsus trajectory summaries normalized by body length.


# Long-format summaries

Instead of manually unpacking list-columns from `res_df`,  call
the convenience wrapper `summarise_kinematics_long()` which returns a
list of long-format tables for common summaries:

```{r summarise-long, eval=FALSE}


kin_long <- summarise_kinematics_long(res)


head(kin_long$leg_times)

head(kin_long$stride_kin)

head(kin_long$tarsus_summary)

head(kin_long$all_measures)


```
## Overview plots

For a quick visual impression of leg-wise kinematics across individuals, legs,
and ant/non-ant grouping the helper `plot_kinematics_overview()`
produces a compact set of diagnostic panels from the combined long-format table
`all_leg_measures`.

```{r kinematics-overview, fig.width=7, fig.height=5, eval=FALSE}
plot_kinematics_overview(kin_long$all_leg_measures)
```
