---
title: "Comparing ants and non-ants with linear mixed models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing ants and non-ants with linear mixed models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

This vignette demonstrates how to analyse stride- and cycle-wise gait
measures using linear mixed-effects models. The main question is
whether ants differ from non-ants in centre-of-mass (COM) stride length,
COM stride speed, and duty factor while accounting for repeated
measurements within individuals and legs.

We assume that kinematic measures have already been computed and merged
into a single long-format tibble `all_leg_measures`. Each row
corresponds to one stride (for spatial measures) or one cycle (for
temporal measures) of a given leg.

```{r packages, eval=FALSE}
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest) 
library(broom.mixed)
library(ggeffects)   
library(patchwork)   
library(emmeans)
library(XML)
library(dplyr)
library(tidyr)
library(broom.mixed)   # or broom.mixed::tidy
library(modelsummary)
library(lmerTest)
library(gtsummary)
library(broom.mixed)  # ensures broom::glance() works nicely for lmer models
library(officer)
library(flextable)
devtools::load_all()
```

# Data 

```{r cleandata, eval=FALSE}

devtools::load_all()
root_data_dir <- "~/Dropbox/hymi/v03/data"

res <- batch_compute_kinematics(
  meta_dir    = root_data_dir,
  meta_file   = "Tabelle_Gesamt_23-4-21.xlsx",
  trk_dir     = root_data_dir,
  plotsize    = 1.5,
  scale_plot  = "bodylength",
  out_fig_dir = "~/tmp/gothouant/"
)

kin_long <- summarise_kinematics_long(res)

d_speed <- kin_long$stride_kin |>
  dplyr::select(c(Individuum, Messnummer,midpoint_stride_speed, body_length, mass_mg, ant_group, Species, Genus)) |>
  na.omit()
str(d_speed)

dat <- kin_long$all_leg_measures |>
  dplyr::select(c(Individuum, Messnummer,metric, value, mass_mg, ant_group, Species,Genus, cycle, leg_num, body_length)) |>
  na.omit()
str(dat)


d_speed <- dat |>
  pivot_wider(names_from = metric, values_from = value) |>
  mutate(contact_inv = 1/contact_time,
         swing_inv = 1/swing_time,
         leg_stride_length_s = scale(leg_stride_length),
         contact_time_s = scale(contact_time),
         swing_time_s = scale(swing_time),
         body_length_s = scale(body_length),
         midpoint_stride_speed_s = scale(midpoint_stride_speed))
d_speed

data_mixed_model <- d_speed

head(data_mixed_model)
usethis::use_data(data_mixed_model, overwrite = TRUE)
writexl::write_xlsx(data_mixed_model, paste0("~/tmp/data_mixed_model-", Sys.Date(), ".xlsx"))

contrasts(d_speed$ant_group)

```

# Mixed Model

## Scaled

```{r modelscaled, eval=FALSE}


m_speed_scaled <- lmer(
  midpoint_stride_speed ~ ant_group * (leg_stride_length_s + contact_time_s + swing_time_s) + body_length_s + (1 | Species) + (1  |Species:Individuum) ,
  data = d_speed
)
summary(m_speed_scaled)

tbl_speed_scaled <- 
  tbl_regression(
    m_speed_scaled,
    exponentiate = FALSE,
    label = list(
      ant_group ~ "Non-ant (vs ant)",
      leg_stride_length_s ~ "Leg stride length (per 1 SD of mm)",
      contact_time_s ~ "Contact time (per 1 SD of s)",
      swing_time_s ~ "Swing time (per 1 SD of s)",
      body_length_s ~ "Body length (per 1 SD of mm)"
    )
  ) |>
  add_glance_table(
    include = c(nobs, AIC, BIC)   # you can also add logLik, deviance, etc.
  )|>
  modify_caption(
    "**Table X. Linear mixed model of midpoint stride speed (mm/s) with standardized predictors.**  
    Fixed-effect estimates represent the expected change in midpoint stride speed (mm/s) associated with a 1 standard deviation increase in each continuous predictor, holding all other variables constant."
  )

doc <- read_docx()
doc <- body_add_flextable(doc, as_flex_table(tbl_speed_scaled))
print(doc, target = "speed_model_scaled.docx")

```

## Unscaled

```{r model, eval=FALSE}

m_speed <- lmer(
  midpoint_stride_speed ~ ant_group * ( leg_stride_length + contact_time + swing_time) + body_length + (1 | Species) + (1  |Species:Individuum) ,
  data = d_speed
)
summary(m_speed)

tbl_speed <- 
  tbl_regression(
    m_speed,
    exponentiate = FALSE,
    label = list(
      ant_group ~ "Non-ant (vs ant)",
      leg_stride_length ~ "Leg stride length (mm)",
      contact_time ~ "Contact time (s)",
      swing_time ~ "Swing time (s)",
      body_length ~ "Body length (mm)"
    )
  ) |>
  add_glance_table(
    include = c(nobs, AIC, BIC)   # you can also add logLik, deviance, etc.
  )|>
  modify_caption(
    "**Table Y. Linear mixed model of midpoint stride speed (mm/s) with predictors in original units.**  
    Fixed-effect estimates represent the expected change in midpoint stride speed (mm/s) per 1-unit increase in each continuous predictor (mm or s), holding all other variables constant."
  )
doc <- read_docx()
doc <- body_add_flextable(doc, as_flex_table(tbl_speed))
print(doc, target = "speed_model_absolute.docx")
```

## Plots

```{r model-plots, eval=FALSE}

caption_stride <- paste(
  "Raw data points (semi-transparent markers) and individual means (opaque marker) illustrate how midpoint speed
varies with leg stride length in this limited sample of ant and non-ant species.
Dashed lines are unadjusted group-wise linear fits. Solid coloured lines and ribbons show mixed-model predictions from a linear
mixed-effects model with random intercepts for species and individuals and an ant_group × leg_stride_length interaction,
adjusted for contact time, swing time, and body length. Thus, the fitted curves reflect the effect of changing
leg stride length while holding the other kinematic predictors and body length constant and averaging over random effects.
Increasing stride length is more strongly associated with higher speed in ants than in non-ants,
indicating that ants benefit more from long steps.")
p_stride <- plot_mixed_effect(
  model = m_speed,
  data = d_speed,
  label_size = 1.5,
  caption = caption_stride,
  title = "Effect of leg stride length on midpoint speed", 
  sizevar = "body_length", 
  xvar = "leg_stride_length",
  ylab = expression("Midpoint speed in "*mm~s^{-1}*""),
  xlab = "Leg stride length in mm",
  size_label = "Body length in mm", 
  add_lit_line = FALSE,
  show_ci = TRUE
)
p_stride
ggsave("speed_stride_length.pdf")


caption_contact <- paste(
  "Raw data points (semi-transparent markers) and individual means (opaque marker) illustrate how midpoint speed
varies with contact time in this limited sample of ant and non-ant species.
Dashed lines are unadjusted group-wise linear fits. Solid coloured lines and ribbons show mixed-model predictions from a linear
mixed-effects model with random intercepts for species and individuals and an ant_group × contact-time interaction,
adjusted for leg stride length, swing time, and body length. Thus, the fitted curves reflect the effect of changing
contact time while holding the other kinematic predictors and body length constant and averaging over random effects.
Decreasing contact time is more strongly associated with higher speed in ants than in non-ants,
consistent with ants regulating speed primarily by shortening stance duration.")
p_contact <- plot_mixed_effect(
  model = m_speed,
  data = d_speed,
  label_size = 1.5,
  caption = caption_contact,
  title = "Effect of contact time on midpoint speed", 
  sizevar = "body_length", 
  xvar = "contact_time",
  ylab = expression("Midpoint speed in "*mm~s^{-1}*""),
  xlab = "Contact time in s",
  size_label = "Body length in mm", 
  add_lit_line = FALSE,
  show_ci = TRUE
)
p_contact
ggsave("speed_contact_time.pdf")


caption_swing <- paste(
  "Raw data points (semi-transparent markers) and individual means (opaque marker) illustrate how midpoint speed
varies with swing time in this limited sample of ant and non-ant species.
Dashed lines are unadjusted group-wise linear fits. Solid coloured lines and ribbons show mixed-model predictions from a linear
mixed-effects model with random intercepts for species and individuals and an ant_group × swing-time interaction,
adjusted for leg stride length, contact time, and body length. Thus, the fitted curves reflect the effect of changing
swing time while holding the other kinematic predictors and body length constant and averaging over random effects.
Longer swing time is clearly more detrimental to speed in ants than in non‑ants, suggesting that ants rely on very rapid leg recovery to maintain their high speeds.")
p_swing <- plot_mixed_effect(
  model = m_speed,
  data = d_speed,
  label_size = 1.5,
  caption = caption_swing,
  title = "Effect of swing time on midpoint speed", 
  sizevar = "body_length", 
  xvar = "swing_time",
  ylab = expression("Midpoint speed in "*mm~s^{-1}*""),
  xlab = "Swing time in s",
  size_label = "Body length in mm", 
  add_lit_line = FALSE,
  show_ci = TRUE
)
p_swing
ggsave("speed_swing_time.pdf")

```


# Tarsi Excursions (Gardner Altman Style Plots)


```{r gardner, eval=FALSE}
kin_long$tarsus_summary
kin_long$leg_times



res_lon_min <- plot_gardner_altman_by_leg(
  data = kin_long$tarsus_summary,
  metric_of_interest = "lon_min",
  subtitle = "Maximum tarsus retraction from midpoint along longitudinal axis",
  ylab = "Retraction (body lengths)",
  ylim_raw = c(-1.2, 1.2),
  ylim_diff = c(-0.6, 0.6)
)
res_lon_min$plot
ggsave("lon_min.pdf", width = 4, height = 7)

res_lon_max <- plot_gardner_altman_by_leg(
  data = kin_long$tarsus_summary,
  metric_of_interest = "lon_max",
  subtitle = "Maximum tarsus protraction from midpoint along longitudinal axis", 
  ylab = "Protraction (body lengths)",
  ylim_raw = c(-1.2, 1.2),
  ylim_diff = c(-0.6, 0.6)
)
res_lon_max$plot
ggsave("lon_max.pdf", width = 4, height = 7)

res_lat_median <- plot_gardner_altman_by_leg(
  data = kin_long$tarsus_summary,
  metric_of_interest = "lat_median",
  subtitle = "Median tarsus–midline distance (mediolateral)", 
  ylab = "Distance (body lengths)",
  ylim_raw = c(0.1, 0.9),
  ylim_diff = c(-0.3, 0.3)
)
res_lat_median$plot
ggsave("lat_median.pdf", width = 4, height = 7)


duty_factor <- kin_long$leg_times |>
  rename(value = duty_factor) |>
  mutate(metric = "duty_factor") |>
  dplyr::filter(!is.na(value))

res_duty <- plot_gardner_altman_by_leg(
  data = duty_factor,
  metric_of_interest = "duty_factor",
  subtitle = "Leg duty factors",
  ylab = "Duty factor",
  ylim_raw = c(0, 100),
  ylim_diff = c(-20, 20)
)

res_duty$plot
ggsave("duty_factor.pdf", width = 4, height = 7)

```

