---
title: "Comparing ants and non-ants with linear mixed models"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Comparing ants and non-ants with linear mixed models}
%\VignetteEngine{knitr::rmarkdown}
\usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

# Introduction

This vignette demonstrates how to analyse stride- and cycle-wise gait
measures using linear mixed-effects models.[cite:1] The main question is
whether ants differ from non-ants in centre-of-mass (COM) stride length,
COM stride speed, and duty factor while accounting for repeated
measurements within individuals and legs.[cite:1]

We assume that kinematic measures have already been computed and merged
into a single long-format tibble `all_leg_measures`.[cite:1] Each row
corresponds to one stride (for spatial measures) or one cycle (for
temporal measures) of a given leg.

```{r packages}
library(dplyr)
library(ggplot2)
library(lme4)
library(lmerTest)  # p-values and convenient summaries
library(broom.mixed)
```

# Data structure

The data used in this vignette should have at least the following
columns:[cite:1]

- `Individuum`: individual ID (character or factor)
- `Messnummer`: measurement or trial number
- `side`: side label (e.g. `"L"` or `"R"`)
- `leg`: leg identifier (e.g. `"L1"`, `"R2"`)
- `ant_group`: factor with levels `"ant"` and `"non-ant"`
- `stride`: stride label (e.g. `"l1"`, `"l2"`)
- `com_stride_length`: COM-based stride length
- `com_stride_speed`: COM-based stride speed
- `duty_factor`: duty factor in percent (from temporal tibble)

```{r glimpse, eval=FALSE}
dplyr::glimpse(all_leg_measures)
```

For the examples below we will restrict to strides with complete data
for the focal outcome and drop obviously invalid values (e.g. negative
speeds).

```{r clean-data}
library(ggeffects)   # ggpredict / ggeffect
library(ggplot2)
library(patchwork)   # or cowplot

root_data_dir <- "~/Dropbox/hymi/v03/data"

res <- batch_compute_kinematics(
  meta_dir    = root_data_dir,
  meta_file   = "Tabelle_Gesamt_23-4-21.xlsx",
  trk_dir     = root_data_dir,
  plotsize    = 1.5,
  scale_plot  = "bodylength",
  out_fig_dir = "~/tmp/gothouant/"
)

kin_long <- summarise_kinematics_long(res)

d_speed <- kin_long$stride_kin |>
  dplyr::select(c(Individuum, Messnummer,midpoint_stride_speed, body_length, mass_mg, ant_group, Species)) |>
  na.omit()
str(d_speed)

dat <- kin_long$all_leg_measures |>
  dplyr::select(c(Individuum, Messnummer,metric, value, mass_mg, ant_group, Species, cycle, leg_num, body_length)) |>
  na.omit()
str(dat)

colnames(dat)

d_speed <- dat |> pivot_wider(names_from = metric, values_from = value)
d_speed

contrasts(d_speed$ant_group)

```

# Example 1: COM stride speed

## Model specification

We start with COM stride speed as outcome and ant-group as the main
predictor.

```{r model-speed}
m_speed_1 <- lmer(
  midpoint_stride_speed ~ ant_group + body_length + (1 | Species) + (1  | Individuum) ,
  data = d_speed
)

m_speed_2 <- lmer(
  midpoint_stride_speed ~ ant_group * (scale(leg_stride_length) + scale(contact_time) + scale(swing_time) + scale(body_length)) + (1 | Species) + (1  |Species:Individuum) ,
  data = d_speed
)

summary(m_speed_2)
BIC(m_speed_2)


# predictions over body_length for each ant_group
pred <- ggpredict(
  m_speed,
  terms = c("body_length", "ant_group")
)

p_raw <- ggplot(d_speed,
                aes(x = body_length,
                    y = midpoint_stride_speed,
                    colour = ant_group)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(title = "Raw data")

p_fit <- plot(pred) +
  theme_minimal() +
  labs(title = "Model predictions")

p_raw + p_fit   # with patchwork
# or cowplot::plot_grid(p_raw, p_fit, nrow = 1)
```

The fixed-effect estimate for `ant_groupant` corresponds to the mean
difference in COM stride speed between ants and non-ants, with
non-ants as the reference level.[cite:1]

For a more compact table:

```{r tidy-speed}
broom.mixed::tidy(m_speed, effects = "fixed", conf.int = TRUE)
```

## Model diagnostics

```{r diag-speed, fig.cap="Residuals vs fitted for COM stride speed model"}
par(mfrow = c(1, 2))
plot(m_speed, which = 1)  # residuals vs fitted
qqnorm(resid(m_speed)); qqline(resid(m_speed))
par(mfrow = c(1, 1))
```

If strong deviations from normality or heteroscedasticity are visible,
consider transforming the outcome (e.g. log-speed) or using a robust
variance estimator.[cite:1]

## Marginal means

We can obtain group means from the fitted model by predicting at the
group level and averaging.[cite:1]

```{r emmeans-speed}
pred_dat <- dat %>%
  group_by(ant_group) %>%
  summarise(
    com_stride_speed = mean(com_stride_speed, na.rm = TRUE),
    .groups = "drop"
  )

pred_dat
```

For more advanced marginal means and contrasts you may integrate
`emmeans`, but this vignette stays with base tools to keep dependencies
minimal.

# Example 2: COM stride length

We repeat the same structure for COM stride length.[cite:1]

```{r model-length}
m_length <- lmer(
  com_stride_length ~ ant_group + (1 | Individuum) + (1 | Individuum:leg),
  data = dat
)

broom.mixed::tidy(m_length, effects = "fixed", conf.int = TRUE)
```

You can add body mass as a covariate if it is available (e.g.
`mass_mg`):[cite:1]

```{r model-length-mass, eval=FALSE}
m_length_mass <- lmer(
  com_stride_length ~ ant_group + scale(mass_mg) +
    (1 | Individuum) + (1 | Individuum:leg),
  data = dat
)

broom.mixed::tidy(m_length_mass, effects = "fixed", conf.int = TRUE)
```

Including mass helps separate taxon-related effects from simple
allometric scaling.[cite:1]

# Example 3: Duty factor

Duty factor is modelled analogously.[cite:1]

```{r model-duty}
m_duty <- lmer(
  duty_factor ~ ant_group + (1 | Individuum) + (1 | Individuum:leg),
  data = dat
)

broom.mixed::tidy(m_duty, effects = "fixed", conf.int = TRUE)
```

Because duty factor is a percentage bounded between 0 and 100, you may
also consider modelling it on the logit scale if values are near the
bounds.[cite:1]

```{r model-duty-logit, eval=FALSE}
dat <- dat %>%
  mutate(
    duty_prop = pmin(pmax(duty_factor / 100, 1e-4), 1 - 1e-4),
    logit_duty = qlogis(duty_prop)
  )

m_duty_logit <- lmer(
  logit_duty ~ ant_group + (1 | Individuum) + (1 | Individuum:leg),
  data = dat
)

broom.mixed::tidy(m_duty_logit, effects = "fixed", conf.int = TRUE)
```

# Random-effects structure

The examples above use random intercepts for individuals and legs within
individuals.[cite:1] This is a pragmatic starting point that captures
the main non-independence structure:

- repeated strides per leg,
- multiple legs per individual.

Depending on sample size, you might also try more complex structures,
such as random slopes of `ant_group` by species or genus, if those are
included in the dataset.[cite:1]

```{r model-speed-random-slopes, eval=FALSE}
m_speed_slopes <- lmer(
  com_stride_speed ~ ant_group +
    (1 + ant_group | Genus) +
    (1 | Individuum) + (1 | Individuum:leg),
  data = dat
)
```

Such models require enough levels per grouping factor to estimate the
variance components reliably.[cite:1]

# Reporting

When reporting results, it is helpful to present:

- estimated mean difference between ants and non-ants,
- 95% confidence intervals,
- approximate p-values (from `lmerTest`),
- sample sizes (number of individuals, number of legs, number of
strides/cycles per group).[cite:1]

A brief summary table for COM stride speed might be constructed as:

```{r summary-report-speed}
tidy_speed <- broom.mixed::tidy(m_speed, effects = "fixed", conf.int = TRUE)

tidy_speed
```

---

This vignette is intentionally minimal and focuses on the workflow:
prepare long-format data, explore distributions, specify a sensible
mixed-effects structure, and extract interpretable contrasts between
ants and non-ants.[cite:1] You can extend it with additional outcomes
(e.g. leg-based stride length), interaction terms (e.g. group × side),
or Bayesian models as needed.

Do you already have `all_leg_measures` saved as an RDS/CSV inside the package, or should the vignette also show how to construct it from the per-leg tibbles?
